---
title: "gapminder extra"
author: "Tim Dennis"
date: "3/30/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## GGPLOT

Let's first load packages we'll need: 

```{r load-gap}
library(tidyverse)
library(gapminder)
library(ggrepel)
```


```{r look-gapminder}
gapminder
```

Show worked example firsts. 

Show simpliest version.

For this exercise, I want to look at only `2007` data. How do we do this in `dplyr`? What function do we use? 

Right. We filter `gapminder` by year. 

```{r filter-gap-year}
gapminder %>% 
  filter(year == "2007")
```

We could save this as a new object `gapminder2007 <- gapminger %>% ...`, but let's pipe the output of the above right into ggplot. Here I want to plot the log of `gdpPercap` against `lifeExp`.

```{r gapminder2007-ggplot}
gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(log(gdpPercap), lifeExp)) +
  geom_point()
```

Let's add color by country to our scaterplot. At the same time, I'll add a `size` and `alpha` attribute to the points. Note, this will obscure our graph b/c there are too many countries and it overwhelms the legend and chart. We'll solve that as we move along. 

```{r gap}
gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(log(gdpPercap), lifeExp)) +
  geom_point(
    aes(color = country),
    size = 3.5,
    alpha = .9) +
  theme_minimal(14) +
  #theme(panel.grid.minor = element_blank()) +
  labs(
    x = "log(GDP per capita)",
    y = "life expectancy")
```

Ok, now we are going to add the country as an annotation to the plot. To do this we are going to use the `ggrepel` package. This package gives a function, `geom_text_repel`, that adds text directly to the plot. It takes an argument for mapping what we want to annotate. Here, we'll add an `aes` of label for country.  

```{r pressure, echo=FALSE}
gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(log(gdpPercap), lifeExp)) +
  geom_point(
    size = 3.5,
    alpha = .9,
    shape = 21,
    col = "white",
    fill = "#0162B2") +
  geom_text_repel(aes(label = country)) +
  theme_minimal(14) +
  theme(panel.grid.minor = element_blank()) +
  labs(
    x = "log(GDP per capita)",
    y = "life expectancy")
```

Still very busy. Next step is to take this approach but sample 10 countries and label the graph with those.


Use `sample()` to select 10 random countries to plot
(run the set.seed() line  rst if you want the same
results)
In the `mutate()` call, check if country is one of the
countries in ten_countries. If it's not, make the label
an empty string (""),
Add the text repel geom from ggrepel. Set the label
aesthetic using the variable just created in `mutate()`

```{r sample10}
library(gapminder)
library(ggrepel)
set.seed(42)

ten_countries <- gapminder$country %>%
  levels() %>%
  sample(10)

ten_countries
```

```{r}
p1 <- gapminder %>%
  filter(year == 2007) %>%
  mutate(label = ifelse(country %in% ten_countries, as.character(country), "")) %>%
  ggplot(aes(log(gdpPercap), lifeExp)) +
  geom_point(size = 3.5,
             alpha = .9,
             shape = 21,
             col = "white",
             fill = "#0162B2"
)
p1
```

```{r}
scatter_plot <- p1 +
  geom_text_repel(aes(label = label), size = 4.5, point.padding = .2, box.padding = .3, force = 1, min.segment.length = 0) +
  theme_minimal(14) +
  theme(legend.position = "none", panel.grid.minor = element_blank()
) +
  labs( x = "log(GDP per capita)", y = "life expectancy")

scatter_plot
```