---
title: "gapminder extra"
author: "Tim Dennis"
date: "3/30/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## GGPLOT

CREDITS: Malcolm Barrett <https://github.com/malcolmbarrett/>

Let's first load packages we'll need: 

```{r load-gap, message=FALSE}
library(tidyverse)
library(gapminder)
library(ggrepel)
```

We'll be using the gapminder data we are familiar with. We can look at the top of that dataframe by piping it to `head()`.

```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(mean_gdpPercap = mean(gdpPercap)) %>% 
  ggplot(aes(x=year, y = mean_gdpPercap, color=continent)) + geom_point()
```

```{r look-gapminder}
gapminder %>% head()
```

Another trick to get the gapminder to show up as an object in your environment in RStudio is to run `data(gapminder)`:

```{r load_gapminder}
data("gapminder")
```

Remember to get help in R we can run a question mark before the function we want help on: 

```{r get-help, eval=FALSE}
?data
```

Let's look at what we will building up to in this episode. It's a scatter plot of countries by GDP per capita and life expectancy for the year 2007. To get a sense of countries in the graph we'll pick 10 countries and annotate the plot with where they are on the graph. We'll color and size the dots and add some labels. Let's see what it looks like:  

```{r sample10_countries, echo=FALSE, message=FALSE}
set.seed(42)

ten_countries <- gapminder$country %>%
  levels() %>%
  sample(10)
```

```{r scatter-plot, echo=FALSE}
gdp_lifeexp <- gapminder %>%
  filter(year == 2007) %>%
  mutate(label = ifelse(country %in% ten_countries, as.character(country), "")) %>%
  ggplot(aes(log(gdpPercap), lifeExp)) +
  geom_point(size = 3.5,
             alpha = .9,
             shape = 21,
             col = "white",
             fill = "#2774AE"
) +
  geom_text_repel(aes(label = label), size = 4.5, point.padding = .2, box.padding = .3, force = 1, min.segment.length = 0) +
  theme_minimal(14) +
  theme(legend.position = "none", panel.grid.minor = element_blank()
) +
  labs( x = "log(GDP per capita)", y = "life expectancy")

gdp_lifeexp
```

Ok, we know what we are aiming for, how do we get there? Any ideas? Intuitions? First steps?

Right. We need to prepare the data first! We'll filter it by the year 2007. Who remember what function we use to filter data?

```{r filter-gap-year}
gapminder %>% 
  filter(year == "2007")
```

We could save this as a new object `gapminder2007 <- gapminger %>% ...`, but let's pipe the output of the above right into ggplot. This is a quick way to explore your data first before saving objects in your environment. Here we'll plot the log of `gdpPercap` against `lifeExp`.

```{r gapminder2007_ggplot}
gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(log(gdpPercap), lifeExp)) +
  geom_point()
```

Looks familiar right! But we have no sense of what countries are plotted and where they are. At this point, we often add a third variable to the plot through color. So we will color our plot by a variable in our data, but this often is problematic b/c it adds mental burden to our plot. Let's show this with countries. 

simplify aesthetics and highlightdesign 

```{r gap}
gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(log(gdpPercap), lifeExp)) +
  geom_point(
    aes(color = country),
    size = 3.5,
    alpha = .9) 
```

The legend overtakes our plot and become really difficult to understand. Instead of overloading our plots we should strive to simplify the aesthetics and when at all possible design without using legends. We can do this with direct labelling. To label the plot we'll use the `ggrepel` package. It give us two functions:  `geom_text_repel()` & `geom_label_repel()`.`geom_text_repel` adds text directly to the plot. It takes an argument for mapping what we want to annotate. Here, we'll add an `aes` of label for country.  We'll also change the size of the dots, set a transparency (alpha), add a color and fill. What does color do? Fill? Yes, colors around the edge of the dots and fill fills the dot with color. We'll use the UCLA blue for the color. 

```{r pressure, echo=FALSE}
gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(log(gdpPercap), lifeExp)) +
  geom_point(
    size = 3.5,
    alpha = .9,
    shape = 21,
    col = "white",
    fill = "#2774AE") +
  geom_text_repel(aes(label = country)) 
```

Still very busy. Next, we want to subset the data by 10 countries to reduce the cognitive load on the graph. We can do this any number of ways, but here we'll use the `sample()` function to randomly select ten countries. We'll also use the `set.seed()` function so it sets our countries to the same 10 countries if we rerun the function. If you want different set of 10 countries each time you run the code, remove or comment out the `set.seed()` line. 

```{r sample10}
set.seed(42)

ten_countries <- gapminder$country %>%
  levels() %>%
  sample(10)

ten_countries
```

Next step is to use the `ten_countries` vector to create a new variable in our dataframe that will become the countries we label the plot with.  So, in the `mutate()` function, we check if  each country is one of the
countries in `ten_countries`. If it's not, make that label row an empty string (""). If it is, we add the country to that row as a character. I'll assign the output of this code to a new called  `gapminder_label`. 


```{r gap_labeled}
gapminder_label <- gapminder %>%
  filter(year == 2007) %>%
  mutate(label = ifelse(country %in% ten_countries, as.character(country), "")) 
```

Let's look at the data. Click Next on the code chunk output in Rmarkdown until you see a country in the `label` column.

```{r}
gapminder_label
```

Ok, we have a new dataframe with a column for the countries we want to add labels for dots on our scatterplot. We can now use this dataframe to label the graph. 

```{r plot_gap_labeled}
gapminder_label %>%   
  ggplot(aes(log(gdpPercap), lifeExp)) +
  geom_point(size = 3.5,
             alpha = .9,
             shape = 21,
             col = "white",
             fill = "#0162B2"
) +
  geom_text_repel(aes(label = label), size = 4.5, point.padding = .2, box.padding = .3, force = 1, min.segment.length = 0) 
```

This is a much easier graph to understand. Let's clean it up a bit by adjusting the grid, theme and labels.

```{r gapminder_label_theme}
gapminder_label %>%   
  ggplot(aes(log(gdpPercap), lifeExp)) +
  geom_point(size = 3.5,
             alpha = .9,
             shape = 21,
             col = "white",
             fill = "#0162B2"
) +
  geom_text_repel(aes(label = label), size = 4.5, point.padding = .2, box.padding = .3, force = 1, min.segment.length = 0) +
  theme_minimal(10) +
  theme(legend.position = "none", panel.grid.minor = element_blank()
) +
  labs( x = "log(GDP per capita)", y = "life expectancy")

```

Ok, next step might be to change the countries. Say you want to specify ten countries manually? How would you go about this? 
